Lazy Linearity for a Core Functional Language

### 2025-10-15 popl26.tex
Almost finish Let Bindings S2

Lastly, consider a (ill-typed) program which defines two let bindings \incode{z} and \incode{y}, where
\incode{z} uses \incode{y} which in turn uses the linear resource \incode{x}:
%
\begin{noway}
\begin{lstlisting}
f2 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* ()
f2 use x = let y = use x in let z = use y in ()
\end{lstlisting}
\end{noway}
%
Even though the binding \incode{y} is used in \incode{z}, \incode{x} is never
consumed because \incode{z} isn't evaluated in the let body, and, thus,
\incode{y} isn't evaluated either. We use this example to highlight that even
for let-bound variables, the syntactic occurrence of a variable isn't enough to
determine whether it is used. Instead, we ought to think of uses of \incode{y}
as implying using \incode{x}, and therefore uses of \incode{z} imply using
\incode{x}, however, if neither are used, then \incode{x} isn't used. Since
\incode{x} is effectively discarded, this example also violates linearity.

>>>>

Lastly, consider an ill-typed program which defines two let bindings \incode{z}
and \incode{y}, where \incode{z} uses \incode{y} which in turn uses the linear
resource \incode{x}:
%
\begin{noway}
\begin{lstlisting}
f2 :: (a *' $\lolli$ '* a) -> a *' $\lolli$ '* ()
f2 use x = let y = use x in let z = use y in ()
\end{lstlisting}
\end{noway}
%
Even though \incode{y} occurs in \incode{z}, \incode{x} is never consumed
\ROUNDTWO{because \incode{z} is discarded}. \ROUNDTWO{This example highlights
that naively checking for syntactic occurrences of let-bound variables such as
\incode{y} is also inadequate to track resource usage}.
%
\ROUNDTWO{A better mental model is that} using \incode{y} implies
using \incode{x}, and thus using \incode{z} implies using \incode{x}
\ROUNDTWO{too}. If neither \incode{z} nor \incode{y} are used, then \incode{x}
is \ROUNDTWO{effectively discarded}.
